Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-25 MScDiplom_Lemenkova.tex
tracked down and corrected misprints

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appretiate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using Gdal technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throuout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am greatful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr.\textit{Dingtian Yang} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectances of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

	\textbf{Software and utilities used in the current work}:\\
	\begin{itemize}
		\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
		\item \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and 					classification
		\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water 					properties under changing environment
		\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and GPS for data capture during the fieldwork
		\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
		\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
		\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
		\item \href{http://www.gnuplot.info/}{Gnuplot} for plotting graphs
		\item \href{http://www.openoffice.org/}{Open Office} : document typesetting, data preprocessing and preliminary statistical evalution
		\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX} : converting spreadsheets into \LaTeX environment
		\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)} : statistical data edition and analysis 
		\item \href{http://gretl.sourceforge.net/}{Gretl: Gnu Regression, Econometrics and Time-series Library} for statistic analysis
		\item \href{http://staruml.sourceforge.net/en/}{StarUML} , \href{http://live.gnome.org/Dia}{Dia} , \href{http://argouml.tigris.org/}						{ArgoUML}: diagramming flowcharts for work algorithms
		\item \href{http://inkscape.org/}{Inkscape}: drawing
		\item \href{http://www.notepad-plus-plus.org/}{Notepad++}: auxiliary text editing and raw data formatting
	\end{itemize}
\pagebreak

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to marine environment, they are however, more closely related to the lily family
(Liliaceae) than to true grass, despite their name “seagrass”, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}.\\
Seagrasses create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as a valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global ditribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determines grouth of seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass Posidonia oceanica (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

and is widely spread along the coasts of Crete \cite{Dumay02} \label{Dumay02}. 
It plays an important role in a number
of geomorphological and ecological processes. Namely, it is a source of food for herbivorous fauna as
well as helter zones for fish and other marine organisms; it contributes to the nutrient recycling; it
provides sediments stability by reducing the degree of water movements, etc \cite{Francour99}\label{Francour99}.\\
The purpose of current MSc research work aims to apply methods of remote sensing and GIS
including processing and classification of satellite, aerial photos and videometric underwater imaging
as well as GIS-based spatial analysis towards mapping and environmental monitoring of \emph{P. oceanica}
along the coasts of Crete island, Greece. The technical implementation is based on ENVI, Erdas
Imagine, WASI and ArcGIS software using aerial and satellite images as well as results of the
underwater videometric measurements.

@@

The distribution of the seagrasses is strongly influensed by several environmental factors, which
include climate (mostly, tropical and temperate are--as), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutritient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: Zostera marina,
Zostera noltii, Cymodocea nodosa and \textit{P.oceanica}. In Greece the common species are\textit{P.oceanica}
(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann and Halophila
stipulacea \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and
phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, Cymodocea nodosa is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When P.
oceanica beds regresses, C. nodosa often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
C. nodosa, and Z. noltii do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

Finally, seagrasses are an important component in the environmental "food chain" of the coastal
ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.\\
Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine “algae”, “seagrasses” are vastly different from them. There are
fundamental differences between both marine organisms, the major of them should be briefly
mentioned: first, seagrasses are true plants with root system and leaves which photosynthesise, have
complex structure and create landscape-similar vast formations on the colonised seafloor areas with
soft sediment, whereas algae are simple organisms that can only holdfast; secondly, seagrasses are
complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while algae
have simple few cell structure with spores and gametes; finally, seagrasses uptake nutrient through
root system, while algae nourish directly from the water column \cite{Dixon05}\label{Dixon05}. There are other
differences between “seagrass” and “algae” but they go beyond the scope of this work.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (P.
oceanica) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commertial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contaminantion and nutrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
This naturally increase environmental pressing and leads to drastical loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.
Other environmental threat for the seagrasses arise as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, partuicularly in areas where P.oceanica is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 percent of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrases on our planet is equal to two footbal
fields per hour \cite{Mellors09b}\label{Mellors09b}.

@@

The dominant and most productive coastal ecosystem of the Mediterranean, P.oceanica is
spatially restricted to the Mediterranean area (Fig.\ref{fig:5}), with its extention limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphological P.oceanica consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

@@

P.oceanica, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:6}): having
usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}).
Growing P.oceanica make a meadows which, in turn, consisnt of smaller patches, called “matte”
(Fig.\ref{fig:7}), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.
Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation

@@

of the meadow health status. \\
There are several environmental factors, determinating the growth of the \textit{P.oceanica}. The adaptation
to dry-summer subtropical climate reduces its extention to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.

@@

The phenology of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in
shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergo
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

\subsection{Research problem}
Monitoring of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlightes
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.\\
Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the sesgrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring.

@@

repeatability and informativeness as a source of data \cite{Dekker05b}\label{Dekker05b}, enabling regular temporal

@@

coverage of the seagrass along the tracklog. Besides, the underwater observations using scuba diving

@@

\subsubsection{Specific objectives}
1. To apply remote sensing data (Google Earth aerial images, Landsat, SPOT satellie images)
for the monitoring of the seagrass meadows distribution
2. To study narrow-band spectral reflectance propoerties of \textit{P.oceanica} and other seafloor
cover types (sand and silt) using WASI swater colour simulation software
3. To use methods of the in situ diving observations and underwater videometric measurements
by Olympus camera in order to receive large-scale imagery of the P.oceanica mattes
4. To perform images classification (supervised and unsupervised) for the thematic mapping of
the P.oceanica seagrass distribution along the coasts of northern Crete.

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.6]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:1.11}
\end{figure}

>>>>

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using Gdal technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr.\textit{Dingtian Yang} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectances of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

	\textbf{Software and utilities used in the current work}:\\

	\vspace{1ex}
	\textbf{Main Tools}:
	\begin{itemize}
		\item \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS 10.0}: mapping, data integration and spatial analysis
		\item \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: imagery processing and 					classification
		\item \href{http://www.filewatcher.com/b/ftp/ftp.dfd.dlr.de/pub/WASI.0.0.html}{WASI Water Color Simulator}, for modelling optical water 					properties under changing environment
		\item \href{http://welcome.hp.com/country/us/en/prodserv/handheld.html}{iPAQ} and GPS for data capture during the fieldwork
		\item \href{http://www.bibtex.org/de/}{BibTeX}, for management of bibliography and reference database
		\item \href{http://www.latex-project.org/}{\LaTeX}, for document typesetting, formatting and performing pdf output
		\item \href{http://www.python.org/}{Python language}, for writing auxiliary scripts (e.g. raw data interpolation)
		\item \href{http://www.gnuplot.info/}{Gnuplot} for plotting graphs
		\item \href{http://www.openoffice.org/}{Open Office} : document typesetting, data pre-processing and preliminary statistical evaluation

		\item \href{http://www.spss.com/}{SPSS (PASW Statistics 18)} : statistical data edition and analysis 
		\item \href{http://gretl.sourceforge.net/}{Gretl: Gnu Regression, Econometrics and Time-series Library} for statistic analysis
		\item \href{http://staruml.sourceforge.net/en/}{StarUML} , \href{http://live.gnome.org/Dia}{Dia} , \href{http://argouml.tigris.org/}						{ArgoUML}: diagramming flowcharts for work algorithms
		\item \href{http://inkscape.org/}{Inkscape}: drawing
	\end{itemize}

	\vspace{1ex}
	\textbf{Auxiliary Tools}:
	\begin{itemize}
		\item \href{http://www.lyx.org/}{LyX} for exporting \LaTeX document to the Word .doc format via HTML
		\item \href{http://calc2latex.sourceforge.net/}{Calc2LaTeX} : converting spreadsheets into \LaTeX environment
		\item \href{http://openev.sourceforge.net/}{OpenEV}: used for raster Google Earth imagery viewing, visualisation and preliminary analysis
		\item \href{http://www.notepad-plus-plus.org/}{Notepad++}: auxiliary text editing and raw data formatting
	\end{itemize}

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to marine environment, they are however, more closely related to the lily family
(Liliaceae) than to the true grass, despite their name “seagrass”, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}.\\
Seagrasses create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass Posidonia oceanica (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

and is widely spread along the coasts of Crete \cite{Dumay02} \label{Dumay02}. 
It plays an important role in a number
of geomorphological and ecological processes. Namely, it is a source of food for herbivorous fauna as
well as shelter zones for fish and other marine organisms; it contributes to the nutrient recycling; it
provides sediments stability by reducing the degree of water movements, etc \cite{Francour99}\label{Francour99}.\\
The purpose of current MSc research work aims to apply methods of remote sensing and GIS
including processing and classification of satellite, aerial photos and videometric underwater imaging
as well as GIS-based spatial analysis towards mapping and environmental monitoring of \emph{P. oceanica}
along the coasts of Crete island, Greece. The technical implementation is based on ENVI, Erdas
Imagine, WASI and ArcGIS software using aerial and satellite images as well as results of the
underwater videometric measurements.

@@

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate are--as), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species in Mediterranean area \cite{Borum04}\label{Borum04}: Zostera marina,
Zostera noltii, Cymodocea nodosa and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann and Halophila
stipulacea \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and
phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, Cymodocea nodosa is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When P.
oceanica beds regresses, C. nodosa often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
C. nodosa, and Z. noltii do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

Finally, seagrasses are an important component in the environmental "food chain" of the coastal
ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.\\
Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine “algae”, “seagrasses” are vastly different from them. There are
fundamental differences between both marine organisms, the major of them should be briefly
mentioned: first, seagrasses are true plants with root system and leaves which photosynthesis, have
complex structure and create landscape-similar vast formations on the colonized seafloor areas with
soft sediment, whereas algae are simple organisms that can only holdfast; secondly, seagrasses are
complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while algae
have simple few cell structure with spores and gametes; finally, seagrasses uptake nutrient through
root system, while algae nourish directly from the water column \cite{Dixon05}\label{Dixon05}. There are other
differences between “seagrass” and “algae” but they go beyond the scope of this work.

@@

The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (P.
oceanica) in the Mediterranean Sea. Nowadays P.oceanica is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
This naturally increase environmental pressing and leads to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subject to catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where P.oceanica is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 percent of the total seagrass meadows have lost any part of their area; the areas, where
the seagrass ecosystems are degrading or lost, are not located in a specific area or continent, but
registered globally; since 1980s global losses of the seagrasses on our planet is equal to two football
fields per hour \cite{Mellors09b}\label{Mellors09b}.

@@

The dominant and most productive coastal ecosystem of the Mediterranean, P.oceanica is
spatially restricted to the Mediterranean area (Fig.\ref{fig:5}), with its extension limited by the western part
of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphological P.oceanica consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots
and short rhizomes (0.5-2.0 mm). The leaves are, perhaps, the most particular characteristics of

@@

P.oceanica, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:6}): having
usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}).
Growing P.oceanica make a meadows which, in turn, consist of smaller patches, called “matte”
(Fig.\ref{fig:7}), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.
Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation

@@

of the meadow health status. \\
There are several environmental factors, determining the growth of the \textit{P.oceanica}. The adaptation
to dry-summer subtropical climate reduces its extension to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.

@@

The phenology of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in
shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergoes 
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

\subsection{Research problem}
Monitoring of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlights 
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.\\
Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the seagrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring.

@@

repeatability and information value as a source of data \cite{Dekker05b}\label{Dekker05b}, enabling regular temporal

@@

coverage of the seagrass along the track log. Besides, the underwater observations using scuba diving

@@

\subsubsection{Specific objectives}
1. To apply remote sensing data (Google Earth aerial images, Landsat, SPOT satellite images)
for the monitoring of the seagrass meadows distribution
2. To study narrow-band spectral reflectance properties of \textit{P.oceanica} and other seafloor
cover types (sand and silt) using WASI water colour simulation software
3. To use methods of the in situ diving observations and underwater videometric measurements
by Olympus camera in order to receive large-scale imagery of the P.oceanica mattes
4. To perform images classification (supervised and unsupervised) for the thematic mapping of
the P.oceanica seagrass distribution along the coasts of northern Crete.

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.57]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:1.11}
\end{figure}
